\rev{Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild}

### 2020-08-07 IJHCSSubmission/bp_results_1.tex
Updates from Overleaf

\section{Observed Trends Over Time in Stress Levels}
\label{stressTrends}

To gain insights into how knowledge workers experience stress over an
extended period of work, we examined the end of day survey responses
collected from each participant to see if any identifiable trends
emerged. We used data from 13 of the 14 participants - we excluded one participant
from this analysis as they experienced atypical stress
levels in the latter half of the study due to factors outside of our
control.

\subsection{Baseline Stress Levels}
Common amongst all participants was a trend to select one stress rating
far more frequently than any other. We will refer to this value as the
participant's baseline stress level. All but one participant reported
their perceived stress level for the day as their baseline stress
level more than 50\% of the time. In total, the baseline values made
up 65\% of the reported values collected from
participants. Interestingly, while participants sometimes saw periods
of sustained increases in stress, lasting as many as 6 consecutive
workdays in the most extreme case, participants would always return to
their baseline stress level at some point.

\begin{figure}
  \centering
      \includegraphics[width=0.7\textwidth]{dailystress_s1.pdf}
      \includegraphics[width=0.7\textwidth]{dailystress_s3.pdf}
  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown. The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed. The x-axis represents the day of the study on which the value was recorded (from 0-45). The x-axes are slightly different for the 2 participants as they did not begin and end the study on exactly the same days.}
   \vspace*{-2mm}
   \label{fig:dailyStress}
\end{figure}

The baseline stress level varied significantly between participants. 7
participants (54\%) reported feeling average stress levels most frequently
(rating 3 on our scale), while 5 (38\%) reported feeling little stress
(rating 2) and 1 (8\%) reported feeling no stress at all (rating 1).

Figure \ref{fig:dailyStress} illustrates these points, showing both participants tendency to report and return to baseline stress levels, as well
as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3).

\subsection{Stressful Days Tend to Cluster}
Accounting for the variance between participants perceived stress
baselines, we consider a stressful day to be one that represents a
deviation of 1 or more stress levels above the participant's
baseline. Of the 93 stressful days we observed in total, we found that
39 (41\%) of these days occurred in groupings of two or more
consecutive stressful workdays. The most common size of these groups
was two workdays, while the largest group we observed was six
workdays.  The day after a stressful day is much more likely to be a
stressful day as compared to any other day with a 0.55 average increase
over baseline, compared to 0.02 average increase over baseline.

\subsection{Extreme Changes in Stress Levels are Rare}
After accounting for each participant's perceived stress baseline, we
examined the frequency of deviations from the baseline. We found that
participants were far more likely to report a stress level that was
within 1 point of their baseline, than to report a stress level 2 or
more points away. These extreme deviations represented only 15\% of
all reported values, which differed from the participants baseline. As
well, the majority (78\%) of these deviations came from just two
participants. This suggests that some people may be less resilient to
the stress of the workplace than others. For most participants,
extremely stressful days were few and far between.

\subsection{Explaining Stress Fluctuations}
In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects. We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables. We also calculated daily features from the gathered computer interaction data where available---namely total time spent actively on the computer, percentage of time idle, and amount of time spent on non work-related web browsing per day. Ultimately, the analysis showed that none of the variables that we examine had a significant explanatory power with respect to our participants perceived stress levels. This indicates that aspects such as the time spent on a computer or the non work-related web browsing do not have a significant impact on the stress level. Furthermore, this points to the need for additional instrumentation and data collection, if we are to successfully understand and predict stress in the workplace.

\section{Predicting Stress in the Moment}
% random forest best
\label{secOverallAccuracy}

To investigate whether stress, focus and awakeness can be predicted in
the moment based on biometric measures,
we investigated classifiers trained
for each individual and across all participants. We report on the effectiveness of
these classifiers and the features that are important in prediction of stress, focus
and awakeness.

@@

In this section we describe how we preprocessed the collected data for use in training and testing machine learning models.

\subsubsection{Data Linking}

We linked the collected biometric data and survey responses for each participant. Linking the data is necessary to construct training and test datasets for use in creating and evaluation machine learning models.

Our linking approach is as follows. From the start time of each survey response, we look back one hour for available biometric data. For each minute in that hour-long time window, we check for biometric data to associate with the survey response. For example, if a participant started a survey response at 11:05am, we look for biometric data in the time frame 10:05am to 11:05am. If biometric data is available in the hour-long time window, we consider the survey response to have associated biometric data. Otherwise, we exclude the survey response from our dataset.

Reasons for a survey response to lack associated biometric data include:
\begin{itemize}
\item The participant not wearing the Everion in the hour before beginning the survey
\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
\item Biometric data not being uploaded successfully to the server
%\item Compatibility issues between the sensor and the OS tracking the participant's data
%\item Issues accessing the data uploaded by the participants
\end{itemize}

@@

From the start time of each survey response, we look back the amount of time 
that corresponds to each time window, and we create features for all of the 
biometric data available in that time window. For example, if a participant 
started a survey response at 11:05am, for the 30min time window, we create 
features using all of the available biometric data from 10:35am to 11:05am. 
For each time window, we calculate 10 statistical measurements from the 
biometric data to create 10 distinct features. Specifically, the 10 
statistical measurements are: mean, standard deviation, variance, median, 
25\textsuperscript{th} percentile, 75\textsuperscript{th} percentile, interquartile range, maximum, minimum, and 
range. Thus, for each survey response, we generate a large number of 
corresponding features based on three factors: biometric measurement, time 
window, and statistical measurement. In addition to these biometric 
features, we also considered the time of day in which the questions were 
asked. These features are created to predict the responses described by the 
ground truth.

\subsubsection{Response Transformations}
Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}). The figure shows that there is a notable imbalance in the distribution of the self-reported responses provided by the participants. Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question. Thus, we binarized the survey data into a two-point scale to give the machine learning models the best possible chance to make useful predictions. The two points in the binary scale represent negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).

We binarized the survey responses as follows. For each participant, we calculated the median response value for each question. We classified each response below the median as 0 ('negative') and each response above the median as 1 ('positive'). The distribution for the stress question skewed left, so we included the median values in the 'positive' class, while the distributions for focus and awakeness skewed right, so we included those median values in the 'negative' class.

\subsubsection{Oversampling}
Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants. This can be seen in the distribution columns in Table \ref{tab:accuracy}. To combat this, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class. This has been a commonly used technique in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.

@@

\begin{table}[h]
	\begin{centering}
	\small\addtolength{\tabcolsep}{-1pt}
    \begin{tabular}{llll}
      \hline
      Variable & K & \# Estimators & Minimum Samples Split \\
      \hline
      Stress & 300 & 100 & 4\\
      Focus & 200 & 50 & 4\\
      Awakeness & 800 & 100 & 4\\
      \hline
    \end{tabular}
    \caption{The hyperparameters selected by grid search analysis to tune our random forest models. K refers to the number of features selected.}    \label{tab:hyperparams}
    \end{centering}
\end{table}

@@

There are many different algorithms that can be used to build a classifier. To select an
algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier. Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network. The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}. For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)

\begin{table*}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
  \caption{Results of predictions using the individual models. The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables. The baseline row represents the averaged results of our baseline classifier. The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
  \vspace*{-4mm}
\end{table*}

\subsection{Individual Classifiers}
%\noindent\textit{Results of Individual Classifiers}\\
Since peoples' experience of stress, focus, and awakeness (as well as
their physiological manifestations) can vary substantially
(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
for each participant (as opposed to a general one for all
participants). The results of our analysis are reported in
Table~\ref{tab:accuracy}. For our analysis, we report values of
accuracy, one of the most commonly used metric to compare performance,
as well as precision and recall of the classes of interest:
'stressed', 'not focused', and 'not awake'. Since the imbalance in the
data can lead to high accuracy values if a classifier always just
predicts the most likely/frequent class while ignoring the class of
higher importance and interest, precision and recall of the class of
interest are also important to
consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
For some users (i.e., S1. as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier. These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.

@@

Overall, we were able to use extracted physiological features to
predict all three aspects with reasonable accuracy, precision, and
recall, as well as to improve upon the baseline---a stratified random
classifier that randomly chooses one of the two classes with a bias
towards the larger class. While the individually trained classifiers
improved on average across all participants upon the baseline in all
cases excepting recall of 'not focused', the improvement was
substantially higher for awakeness (85\% improvement in precision,
62\% in recall, and 7\% in accuracy) than for stress or focus. In addition,
the performance of the individually trained classifiers varied greatly
across participants. While some participants showed a large
improvement, for others the baseline performed much better than the
individually trained classifier. For instance, for predicting
'stressed', the individual classifiers improved upon the baseline for
S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in
precision and 184.6\% in recall for S12, while they did worse for S1,
S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
predict a single instance of 'stressed'. Typically, users that have the
lowest precision and recall values are those where the data is the
most unbalanced.\\[-0.1cm]
% We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
%
%(improvement: 8\% precision, 1\% recall, 8\% accuracy), the individual performance varied greatly among participants. Some participants showed negligible difference in comparison to the baseline classifier, e.g. subject ..., while others showed a large improvement, e.g. subject ...
%\begin{table}[h!]
%\begin{centering}
%\begin{tabular}{lll}
%\hline
%Participant & Recall Improvement (\%) & Precision Improvement (\%)\\
%\hline
%S12 & 184.6 & 88.4 \\
%S6 & 66.7 & 85.8 \\
%S11 & 50.0 & 44.4 \\
%S8 & 32.4 & 26.7 \\
%S14 & 14.7 & 9.8 \\
%S4 & 5.6 & 5.5 \\
%S9 & -55.4 & -46.8 \\
%S3 & -90.0 & -82.1 \\
%S1 & -100.0 & -100.0 \\
%S5 & -100.0 & -100.0 \\
%S7 & -100.0 & -100.0 \\
%S10 & -100.0 & -100.0 \\
%S13 & -100.0 & -100.0\\
%S2 & - & -\\
%\hline
%\end{tabular}
%\caption{Percentage improvement in recall and precision for stress, using our approach compared to the baseline, on an individual level. For participant 2, the baseline achieved a precision and recall of 0, thus the improvement is undefined}
%\end{centering}
%\label{tab:indImprovement}
%\end{table}

\subsection{Feature Selection and Importance}
%\noindent\textit{Feature Selection and Importance}\\
There are a large variety of features that can be (and have been) calculated in previous research for each of the basic measurements listed in Table~\ref{signals}, such as the mean, standard deviation, maximum, and interquartile range. In addition, each of these metrics can be combined with the various time windows captured of a basic measurement, resulting in a large feature space. To reduce the feature space, we experimented with multiple feature selection methods, including selecting the top k highest correlated features by various metrics such as mutual information, Pearson's correlation coefficient, ANOVA's F-value, as well as wrapper methods such as recursive feature elimination, optimizing mean decrease accuracy by iteratively permuting features, and only selecting features that exceed a certain Gini importance threshold. We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models. Ultimately, we chose to use the top k features with the highest ANOVA F-value, as it is relatively simple and efficient to calculate. The values of k used were selected by grid search analysis, and are shown in Table \ref{tab:hyperparams}.

%Respiration rate is also highly important for stress (#3 , 14.4%) but I'm not sure what previous works say about correlation with stress
Overall, the features that were selected as the important ones for the individual models based on the random forest algorithm varied greatly across participants. Yet, some feature categories were considered to be important more frequently than others. Table~\ref{tab:featureImportance} shows the averaged Gini importance for the feature categories used for predicting stress. Particularly important for stress were the feature categories heart rate variability (18.3\%) and skin temperature (15.7\%), both of which have been shown in several previous studies to be  indicators for stress~\cite{dishman2000stress,mcduff16,kataoka00}. For predicting focus, the feature categories for blood pulse wave (14.2\%) and heart rate variability (13\%) showed to be the most important categories, while for awakeness the most important ones were heart rate variability (13.6\%) and blood pulse wave (13.1\%).\\[-0.1cm]

@@

The bottom row of Table \ref{tab:accuracy} presents the averaged performance results for this approach in terms of accuracy, precision, and recall. Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.  Upon closer investigation into the performance of the general model when testing on each participant, we found that individually trained models for each participant performed much better than a general model trained over all participants. Using stress as an example, for participant S12, for whom we saw the greatest increase compared to the baseline in individual models, the general model was unable to predict a single instance of 'stressed' correctly. This is consistent with our expectations because biometric features are highly specific to individuals.

>>>>

\section{Observed Trends Over Time in Stress \rev{and Awakeness} Levels}
\label{stressTrends}

To gain insights into how knowledge workers experience stress \rev{and awakeness} over an
extended period of work, we examined the end of day survey responses
collected from each participant to see if any identifiable trends
emerged. \rev{As we noted in the last section, we did not ask
participants about focus in the end of the day surveys as focus is an 
aspect relevant at a particular moment in time rather than an aspect
for an extended period of work.} We used data from 13 of the 14 participants - we excluded one participant
from this analysis as they experienced atypical stress
levels in the latter half of the study due to factors outside of our
control.

\subsection{Stress Levels}

\rev{We describe characteristics noted in the reporting of the knowledge workers
about the stress they experienced over the duration of the study.}

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.7\textwidth]{s1_stress.pdf}
      \includegraphics[width=0.7\textwidth]{s3_stress.pdf}
  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown. The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed. The x-axis represents the day of the study on which the value was recorded (from 0-45). \rev{Some gaps are present in the chart for S1 as they did not report their stress level on those days.}}
   \vspace*{-2mm}
   \label{fig:dailyStress}
\end{figure}

\subsubsection{Baseline Stress Levels}
Common amongst all participants was a trend to select one stress rating
far more frequently than any other. We will refer to this value as the
participant's baseline stress level. All but one participant reported
their perceived stress level for the day as their baseline stress
level more than 50\% of the time. In total, the baseline values made
up 65\% of the reported values collected from
participants. Interestingly, while participants sometimes saw periods
of sustained increases in stress, lasting as many as 6 consecutive
workdays in the most extreme case, participants would always return to
their baseline stress level at some point.

The baseline stress level varied significantly between participants. 7
participants (54\%) reported feeling average stress levels most frequently
(rating 3 on our scale), while 5 (38\%) reported feeling little stress
(rating 2) and 1 (8\%) reported feeling no stress at all (rating 1). Figure \ref{fig:dailyStress} illustrates these points using \rev{ data from two participants, showing the tendency to report and return to baseline stress levels, as well
as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3).}

\subsubsection{Stressful Days Tend to Cluster}
Accounting for the variance between participants perceived stress
baselines, we consider a stressful day to be one that represents a
deviation of 1 or more stress levels above the participant's
baseline. Of the 93 stressful days we observed in total, we found that
39 (41\%) of these days occurred in groupings of two or more
consecutive stressful workdays. The most common size of these groups
was two workdays, while the largest group we observed was six
workdays.  The day after a stressful day is much more likely to be a
stressful day as compared to any other day with a 0.55 average increase
over baseline, compared to 0.02 average increase over baseline.

\subsubsection{Extreme Changes in Stress Levels are Rare}
After accounting for each participant's perceived stress baseline, we
examined the frequency of deviations from the baseline. We found that
participants were far more likely to report a stress level that was
within 1 point of their baseline, than to report a stress level 2 or
more points away. These extreme deviations represented only 15\% of
all reported values, which differed from the participants baseline. As
well, the majority (78\%) of these deviations came from just two
participants. This suggests that some people may be less resilient to
the stress of the workplace than others. For most participants,
extremely stressful days were few and far between.

\subsection{Awakeness}
\rev{We applied the same analyses described above to the self reported awakeness levels of our participants. Compared with the reported stress levels, we observed the same trend of participants reporting one baseline far more commonly than any other. We did not find there to be a significant correlation between reported stress and awakeness levels. Overall, participants awakeness levels fluctuated significantly less than their stress levels (73.1\% of reports were at the baseline level, compared to 65.1\% for stress, p < 0.05). Participants were unlikely to experience days with heightened (above baseline) awakeness. Such days made up only 6.0\% of the total observed work days across all participant. Most of these days came from one participant, S2, who reported 15 heightened awakeness days compared to the next highest, S13 with 4. Large deviations (>1 point deviation) from each participants baseline stress levels were extremely uncommon, accounting for only 9 (2.1\%) of our total observations. Similarly to what we observed with high stress days, low awakeness days frequently came in clusters of two or three days in a row. Given that the immediately preceding day was a low awakeness day, a given day was 228.2\% more likely to be a low awakeness day than when considering any day at random (p < 0.0001).}

\begin{table}[]
    \centering
    \begin{tabular}{|c|c c|c c|}
    \hline
        &
        \multicolumn{2}{c|}{Stress} & 
        \multicolumn{2}{c|}{Awakeness} \\
        \hline
        \textbf{Explanatory Variable} & $\mathbf{R^{2}}$ & \textbf{p-value} &  $\mathbf{R^{2}}$ & \textbf{p-value}\\
        \hline
        Day of Week - Monday & -0.033 & 0.745 & -0.058 & 0.0.745\\
        Day of Week - Tuesday & 0.028 & 0.773 & 0.036 & 0.662 \\
        Day of Week - Wednesday & 0.163 & 0.100 & -0.018 & 0.830 \\
        Day of Week - Thursday & 0.022 & 0.821 & 0.076 & 0.927 \\
        Day of Week - Friday & 0.082 & 0.473 & -0.178 & 0.025 \\
        Proximity to Month End & 0.044 & 0.687 & -0.082 & 0.374 \\
        \hline
    \end{tabular}
    \caption{Correlation coefficient and associated p-values for the explanatory variables we examined in our linear mixed model analysis.}
    \label{tab:mixedModel}
\end{table}

\subsection{Explaining Fluctuations}
In an attempt to explain some of the \rev{fluctuations in stress and awakeness} that our participants were experiencing, we created a linear mixed model with the self-reported daily stress \rev{and awakeness} levels as dependent variables and the participants as random effects. We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables. Ultimately, the analysis showed that none of the variables that we examined had a significant explanatory power with respect to our participants perceived stress levels.
\rev{For awakeness, we found  that there was a small ($R^2$: -0.178) yet significant decrease in awakeness levels on Fridays in particular. Table \ref{tab:mixedModel} shows some of the detailed results of these analyses.} \rev{These results show the difficulty of explaining a person's stress and awakeness via simplistic measures, and point to the need for additional instrumentation and data collection if we are to successfully understand and make predictions about these human aspects in the workplace.}

\section{\rev{Predicting Stress, Focus and Awakeness in the Moment}}
% random forest best
\label{secOverallAccuracy}

To investigate whether stress, focus and awakeness can be predicted in
the moment based on biometric measures,
we investigated classifiers trained
for each individual and across all participants. We report on the effectiveness of
these classifiers and the features that are important in predicting stress, focus
and awakeness.

@@

\gcm{To ready the collected data for use in training and testing machine learning models, we performed several steps.}

\rev{
\subsubsection{Data Cleaning}
All data recorded by the Everion is associated with a quality score ranging from 0-1 that is calculated using proprietary methods. In accordance with the recommendations of Biovotion,
to prevent our results from being effected by erroneous data we set a quality threshold of 0.5 and discarded any data gathered which had a quality rating below this threshold. 
}
\subsubsection{Data Linking}

We linked the collected biometric data and survey responses for each participant. Linking the data is necessary to construct training and test datasets for use in creating and evaluating machine learning models.

\rev{To link the data, we look back one hour from the start time of each survey response for
available biometric data.  For example, if a participant started a survey response at 11:05am, we look for biometric data from between 10:05am to 11:05am. If no biometric data was recorded in the hour time window, we exclude the survey response from the dataset. Otherwise, we consider the survey response to have associated biometric data.}

There are several reasons for a survey response to lack associated biometric data:
\begin{itemize}
\item The participant was not wearing the Everion in the hour before beginning the survey.
\item The Everion was not recording data in the hour before the participant began the survey (e.g., due to low battery).
\item Biometric data was not being uploaded successfully to the server.
%\item Compatibility issues between the sensor and the OS tracking the participant's data
%\item Issues accessing the data uploaded by the participants
\end{itemize}

@@

From the start time of each survey response, we look back the amount of time 
that corresponds to each time window and we create features for all of the 
biometric data available in that time window. For example, if a participant 
started a survey response at 11:05am, for the 30min time window, we create 
features using all of the available biometric data from 10:35am to 11:05am. 
\rev{If there is a large portion of data missing (either because of a recording issue or because of low quality data) from the time window considered, then the time window is marked as missing. In this case, features are imputed based on the mean of other samples of the same feature for that participant. This is an effective and commonly used technique, which is preferable to the alternative of deletion as it preserves our already small sample size~\cite{hawthorne2005imputing}.}
For each time window, we calculate 10 statistical measurements from the 
biometric data to create 10 distinct features. Specifically, the 10 
statistical measurements are: mean, standard deviation, variance, median, 
25\textsuperscript{th} percentile, 75\textsuperscript{th} percentile, interquartile range, maximum, minimum, and 
range. Thus, for each survey response, we generate a large number of 
corresponding features based on three factors: biometric measurement, time 
window, and statistical measurement. In addition to these biometric 
features, we also considered the time of day in which the questions were 
asked. These features are created to predict the responses described by the 
ground truth. \rev{To attempt to account for inter-participant differences, we normalized all features on a per-participant basis.}

\subsubsection{Response Transformations}
Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}). The figure shows that there is a notable imbalance in the distribution of the self-reported responses provided by the participants. Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question. \rev{Given 
this distribution and based on our earlier observation that the participants tended to adhere to a baseline reporting level for stress and awakeness, we elected to simplify the problem from 5 classes to 2. This transformation enables us to more easily represent patterns in the data, such as when a participant fluctuates from a normal to high stress level. To perform this transformation, we began for
each participant by calculating the median response value for each question. We classified each response below the median as 0 ('negative') and each response above the median as 1 ('positive'). The distribution for the stress question skewed left, so we included the median values in the 'positive' class, while the distributions for focus and awakeness skewed right, so we included those median values in the 'negative' class.}

With this method, we transformed the survey data into a two-point scale, representing  negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).

\subsubsection{Oversampling}
Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants. This can be seen in the distribution columns in Table \ref{tab:accuracy}. To mitigate this effect, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class. This technique has commonly been used  in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.

@@

\begin{comment}
\begin{table}[h]
	\begin{centering}
	\small\addtolength{\tabcolsep}{-1pt}
    \begin{tabular}{llll}
      \hline
      Variable & K & \# Estimators & Minimum Samples Split \\
      \hline
      Stress & 300 & 100 & 4\\
      Focus & 200 & 50 & 4\\
      Awakeness & 800 & 100 & 4\\
      \hline
    \end{tabular}
    \caption{The hyperparameters selected by grid search analysis to tune our random forest models. K refers to the number of features selected.}    \label{tab:hyperparams}
    \end{centering}
\end{table}
\end{comment}

@@

There are many different algorithms that can be used to build a classifier. To select an
algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11},  \rev{evaluating each one by using leave-one-out cross validation}. Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network. For the remainder of this paper, \rev{we refer to a random forest classifier}.%\\[-0.1cm]
% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)

\begin{table*}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{rq1performance_v2.pdf}
  \caption{Results of predictions using the individual models. \rev{The distribution columns show the proportion of the minority class out of the total number of responses for each of the three variables.} The baseline rows represents the averaged results of our baseline classifiers. The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
  \vspace*{-4mm}
\end{table*}

\subsection{Individual Classifiers}
%\noindent\textit{Results of Individual Classifiers}\\
Since peoples' experience of stress, focus, and awakeness (as well as
their physiological manifestations) can vary substantially
(e.g.,~\cite{Hernandez11}), we first trained \rev{and evaluated individual classifiers
for each participant (as opposed to a general one for all
participants) using leave-one-out cross validation.} The results of our analysis are reported in
Table~\ref{tab:accuracy}. For our analysis, we report values of
accuracy, one of the most commonly used metric to compare performance,
as well as precision and recall of the classes of interest:
'stressed', 'not focused', and 'not awake'. Since the imbalance in the
data can lead to high accuracy values if a classifier always just
predicts the most likely/frequent class while ignoring the class of
higher importance and interest, precision and recall of the class of
interest are also important to
consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.

\rev{Beside our results, for the purposes of comparison we also present two commonly used baselines in this table - a majority classifier, which always predicts the larger of the two classes considered, and a stratified random classifier which randomly chooses between the two classes, but with a proportional bias towards the larger class.}

For some users (i.e., S1. as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier. These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.

@@

Overall, we were able to use extracted physiological features to
predict all three aspects with reasonable accuracy, precision, and
recall. \rev{We present a comparison between the averaged results of our individual classifiers and those of the baseline stratified random classifer in the ``Improvement Over Random'' row of Table \ref{tab:accuracy}. This is calculated as the difference between the overall average and the baseline results, divided by the baseline results (for example, $\frac{Acc_{Overall} - Acc_{Random}}{Acc_{Random}}$). We do not compare our results with the majority classifier directly as this classifier achieved precision and recall scores of zero, making a meaningful comparison impossible. The improvement percentages demonstrate that the predictions made by our classifiers are much better than random after correcting for the imbalance in our dataset.}

While the individually trained classifiers
improved on average across all participants upon the baseline in all
cases excepting recall of 'not focused', the improvement was
substantially higher for awakeness (\rev{52.4\%} improvement in precision,
\rev{40.8\%} in recall, and \rev{6.5\%} in accuracy) than for stress or focus. In addition,
the performance of the individually trained classifiers varied greatly
across participants. While some participants showed a large
improvement, for others the baseline performed much better than the
individually trained classifier. For instance, for predicting
'stressed', the individual classifiers improved upon the baseline for
S4, S6, S8, S11, S12, and S14 with a maximum improvement of \rev{152.0\%} in
precision and \rev{111.1}\% in recall for S12, while they did worse for S1,
S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
predict a single instance of 'stressed'. Typically, users that have the
lowest precision and recall values are those where the data is the
most unbalanced.\\[-0.1cm]
% We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
%
%(improvement: 8\% precision, 1\% recall, 8\% accuracy), the individual performance varied greatly among participants. Some participants showed negligible difference in comparison to the baseline classifier, e.g. subject ..., while others showed a large improvement, e.g. subject ...
%\begin{table}[h!]
%\begin{centering}
%\begin{tabular}{lll}
%\hline
%Participant & Recall Improvement (\%) & Precision Improvement (\%)\\
%\hline
%S12 & 184.6 & 88.4 \\
%S6 & 66.7 & 85.8 \\
%S11 & 50.0 & 44.4 \\
%S8 & 32.4 & 26.7 \\
%S14 & 14.7 & 9.8 \\
%S4 & 5.6 & 5.5 \\
%S9 & -55.4 & -46.8 \\
%S3 & -90.0 & -82.1 \\
%S1 & -100.0 & -100.0 \\
%S5 & -100.0 & -100.0 \\
%S7 & -100.0 & -100.0 \\
%S10 & -100.0 & -100.0 \\
%S13 & -100.0 & -100.0\\
%S2 & - & -\\
%\hline
%\end{tabular}
%\caption{Percentage improvement in recall and precision for stress, using our approach compared to the baseline, on an individual level. For participant 2, the baseline achieved a precision and recall of 0, thus the improvement is undefined}
%\end{centering}
%\label{tab:indImprovement}
%\end{table}

\subsection{Feature Selection and Importance}
%\noindent\textit{Feature Selection and Importance}\\
There are a large variety of features that can be (and have been) calculated in previous research for each of the basic measurements listed in Table~\ref{signals}, such as the mean, standard deviation, maximum, and interquartile range. In addition, each of these metrics can be combined with the various time windows captured of a basic measurement, resulting in a large feature space. To reduce the feature space, we experimented with multiple feature selection methods, including selecting the top k highest correlated features by various metrics such as mutual information, Pearson's correlation coefficient, ANOVA's F-value, as well as wrapper methods such as recursive feature elimination, optimizing mean decrease accuracy by iteratively permuting features, and only selecting features that exceed a certain Gini importance threshold. We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models. Ultimately, we \rev{elected not to utilize any feature selection in order to simplify our approach, as we found there to be minimal differences in performance between the techniques, and the random forest algorithm is capable of handling datasets with many features.}

%Respiration rate is also highly important for stress (#3 , 14.4%) but I'm not sure what previous works say about correlation with stress
Overall, the features that were selected as the important ones for the individual models based on the random forest algorithm varied greatly across participants. Yet, some feature categories were considered to be important more frequently than others. Table~\ref{tab:featureImportance} shows the averaged Gini importance for the feature categories used for predicting stress. \rev{Heart rate variability proved to be an important measure for all of the aspects of interest, ranking as the most important feature category for both stress and awakeness, and the second most important one for focus. This is not surprising, as heart rate variability and skin temperature have been shown in several previous studies to be possible indicators for stress levels~\cite{dishman2000stress,mcduff16,kataoka00}. We also found blood pulse wave to be an important indicator for both focus and awakeness, but less important for stress, while respiration rate was important in stress and focus but not awakeness. Besides these mentioned feature categories, there was great variation in which measures were important to which aspect. This shows that there is a clear benefit to having multiple biometric streams available for predicting these aspects.}\\[-0.1cm]

@@

The bottom \rev{3 rows} of Table \ref{tab:accuracy} present the averaged performance results for this approach in terms of accuracy, precision, and recall\rev{, as well as the results of baseline stratified random and majority classifiers following the same leave-one-group-out cross-validation procedure.} Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.  Upon closer investigation into the performance of the general model when testing on each participant, we found that individually trained models for each participant performed much better than a general model trained over all participants. Using stress as an example, for participant S12, for whom we saw the greatest increase compared to the baseline in individual models, the general model was unable to predict a single instance of 'stressed' correctly. This is consistent with our expectations because biometric features are highly specific to individuals.
